Multidimensional Persistence Module Classification via Lattice-Theoretic Convolutions

### 2020-11-26 paper/paper.tex
minor edits new figs

\subsection*{Related work}

Very recent advances have been made to featurize multidimensional persistence modules via
landscapes \cite{vipond2020multiparameter} and images \cite{carriere2020multiparameter}.
Our naive approach at featurizing $2$-dimensional persistence modules has the advantage
of being readily computable with existing software \cite{lesnick_interactive_2015}.
Persistence modules supported on lattices are of increased interest.
By leveraging M\"{o}bius inversion, these persistence modules are shown to generate a stable persistence diagram \cite{mccleary2020edit} as well as factor through a nice category with pseudo-inverses \cite{krishnan2020invertibility}.

@@

Traditional convolutional neural networks are useful in part because the
convolution kernels (here the functions $g$) can have very small support,
reducing the number of parameters that must be learned. In the standard
convolutional setting, these kernels are implicitly supported in a neighborhood
of the origin, but the location of the kernel is not usually explicitly
specified. In the lattice setting, we do need to specify where the kernel
resides. In the abelian group case, the kernel is supported near the identity,
and similarly, when we treat our domain as a lattice, the kernel should be
supported near the neutral element of the operation. That is, for a meet
convolution, $g$ should be supported at the maximum $(m,n)$, and for a join
convolution, $g$ should be supported at the minimum $(0,0)$. This ensures that
the convolution operators are capable of preserving information at every point
of the space. For instance, if $g(0,0) \neq 0$, then $(f \ast_{\vee} g)(x,y)$ is
a sum of terms including $f(x,y)g(0,0)$, so all information from one layer can
be passed to the next layer. 
\section{Experiments}
We use a small portion of the Princeton ModelNet
dataset~\cite{zhirong_wu_3d_2015} as a source of finite
metric spaces. This dataset consists of hundreds of 3-dimensional CAD models
representing objects from 10 classes. We select two of the classes and sample
points from the 3D models to produce finite metric spaces embedded in $\R^3$.
We then compute the corresponding multidimensional persistence modules, from
which we produce features used as an input to a convolutional neural net
classifier.

The pipline thus begins with a 3D polyhedral model, of which $3000$ vertices are
sampled to produce a point cloud in $\R^3$. This point cloud then produces a
bifiltered simplicial complex, whose degree-$0$ persistent homology we calculate
using RIVET~\cite{lesnick_interactive_2015}, sampled at a discrete grid of $40
\times 40$ points, producing lattice-indexed signals given by the Hilbert
function and the multi-graded Betti numbers $\xi_0, \xi_1, \xi_2$; four features
in total. These are then passed to the classifier, which produces a class
prediction, in this case either ``sofa'' or ``monitor.''

As the filter function on these data sets, we use the \introduce{codensity} function \[\rho_{\text{codense}}(x;k) = \left( \frac{1}{k} \sum_{y \in N_k(x)} d(x,y) \right)^{-1},\]
where $N_k(x)$ is the set of the $k$ nearest neighbors to $x$; we select
$k=100$. The name codensity is appropriate because the points in the densest
regions of $\mathcal M$ appear earlier in the filtration. A folk theorem is that
the two-parameter persistent homology of a Rips/codensity bifiltration is stable
under non-Hausdorff perturbations: \textit{the (Rips) persistent homology of a
point sample and another obtained by adding a small number of points at random
are close with respect to the interleaving distance}.

We compare the performance of two convolutional networks on this classification
task. One uses the lattice-convolution based layers described in
Section~\ref{sec:latticeCNN}, and the other uses standard convolutional layers.
Each has three convolutional layers followed by three fully connected layers.
The lattice-based convolution layers are of the form $\alpha\text{MeetConv}(x) +
(1-\alpha)\text{JoinConv}(x)$ for a hyperparameter $\alpha \in [0,1]$. All
convolution kernels have dimension $4 \times 4$, hidden convolution layers have
16 features, and the final convolution has 8 features. The fully connected
layers have 32 features at each inner layer. We use a cross entropy loss
function with a softmax in the final layer. The networks are trained using the
Adam gradient algorithm with learning rate $0.001$ for 50 epochs. We reserve
10\% of the data for testing. Results are shown in Figure~\ref{fig:comparison}.

% The lattice-based convolution and classical convolution networks perform
% similarly on the training set. However, the lattice convolution network's
% predictions generalize more accurately to the testing set ($\approx 7\%$ improvement in testing accuracy).
% This improvement is  at the expense of more variability in performance during
% training. (Our code is available at \texttt{https://github.com/alfred-tarski/latticeconv} .)
\begin{figure}[h]
  \caption{Comparison training curves for the lattice neural network and
    standard convolutional neural network.}\label{fig:comparison}
    \begin{center}
%    \includegraphics[width=0.49\textwidth]{testing-accuracy.png}
%    \vspace{0.2cm}
%    \includegraphics[width=0.49\textwidth]{training-accuracy.png}
%	\includegraphics[width=0.85\textwidth]{figure.png}
    \end{center}
\end{figure}
\section{Discussion}
Our proposed featurization for persistence modules is rather naive, but well
adapted to the use of lattice convolutions as a data processing method. The
lattice convolutional neural network shows promise as a method for classifying
features arising from a multiparameter persistence module. The algebraic
perspective on partially ordered sets, exemplified by lattices, may also offer
approaches to featurizing more complex invariants of persistence modules. In
particular, the incidence algebra may offer a natural way to represent the rank
invariant~\cite{carlsson_theory_2009} in a way amenable to convolution-like
operations.
We hope with these brief experiments to inspire further work on featurizing
multidimensional persistence for use in machine learning algorithms.

\newpage
\small
\bibliographystyle{plain}
\bibliography{latticenn}

>>>>

\paragraph{Related work}
Very recent advances have been made to featurize multidimensional persistence modules via
landscapes \cite{vipond2020multiparameter} and images \cite{carriere2020multiparameter}.
Our naive approach at featurizing $2$-dimensional persistence modules has the advantage
of being readily computable with existing software \cite{lesnick_interactive_2015}.
Persistence modules supported on lattices are of increased interest.
By leveraging M\"{o}bius inversion, these persistence modules are shown to generate a stable persistence diagram \cite{mccleary2020edit} as well as factor through a convenient category with pseudo-inverses \cite{krishnan2020invertibility}.

@@

\paragraph{Remark}
Traditional convolutional neural networks are useful in part because the
convolution kernels (here the functions $g$) can have very small support,
reducing the number of parameters that must be learned. In the standard
convolutional setting, these kernels are implicitly supported in a neighborhood
of the origin, but the location of the kernel is not usually explicitly
specified. In the lattice setting, we do need to specify where the kernel
resides. In the abelian group case, the kernel is supported near the identity,
and similarly, when we treat our domain as a lattice, the kernel should be
supported near the neutral element of the operation. That is, for a meet
convolution, $g$ should be supported at the maximum $(m,n)$, and for a join
convolution, $g$ should be supported at the minimum $(0,0)$. This ensures that
the convolution operators are capable of preserving information at every point
of the space. For instance, if $g(0,0) \neq 0$, then $(f \ast_{\vee} g)(x,y)$ is
a sum of terms including $f(x,y)g(0,0)$, so all information from one layer can
be passed to the next layer. 
\section{Experiments}
We use a large portion of the Princeton ModelNet
dataset~\cite{zhirong_wu_3d_2015} as a source of finite
metric spaces. This dataset consists of hundreds of 3-dimensional CAD models
representing objects from 10 classes. We select two of the classes and sample
points from the 3D models to produce finite metric spaces embedded in $\R^3$.
We then compute the corresponding multidimensional persistence modules, from
which we produce features used as an input to a convolutional neural net
classifier.

The pipline thus begins with a 3D polyhedral model, of which $3000$ vertices are
sampled to produce a point cloud in $\R^3$. This point cloud then produces a
bifiltered simplicial complex, whose degree-$0$ persistent homology we calculate
using RIVET~\cite{lesnick_interactive_2015}, sampled at a discrete grid of $40
\times 40$ points, producing lattice-indexed signals given by the Hilbert
function and the multi-graded Betti numbers $\xi_0, \xi_1, \xi_2$; four features
in total. These are then passed to the classifier, which produces a class
prediction, in this case one of 10 possible household objects.

As the filter function on these data sets, we use the \introduce{codensity} function \[\rho_{\text{codense}}(x;k) = \left( \frac{1}{k} \sum_{y \in N_k(x)} d(x,y) \right)^{-1},\]
where $N_k(x)$ is the set of the $k$ nearest neighbors to $x$; we select
$k=100$.
\paragraph{Remark}
The name codensity is appropriate because the points in the densest
regions of $\mathcal M$ appear earlier in the filtration. A folk theorem is that
the 2-parameter persistent homology of a Rips/codensity bifiltration is stable (w.r.t.~interleaving distance)
under non-Hausdorff noise (e.g. adding or removing a small number of point samples).
%: \textit{the (Rips) persistent homology of a
%point sample and another obtained by adding a (relatively) small number of points at random
%are close with respect to the interleaving distance}.

We compare the performance of two convolutional networks on this classification
task. One uses the lattice-convolution based layers described in
Section~\ref{sec:latticeCNN}, and the other uses standard convolutional layers.
Each has three convolutional layers followed by two fully connected layers.
The lattice-based convolution layers are of the form $\alpha \cdot \text{MeetConv}(x) +
(1-\alpha)\cdot\text{JoinConv}(x)$ for a hyperparameter $\alpha \in [0,1]$. All
convolution kernels have dimension $4 \times 4$, hidden convolution layers have
16 features, and the final convolution layer has 8 features. The first two convolution
layers are followed by max-pooling layers with a $2 \times 2$ kernel.
The inner fully connected layer has 32 features. We use a cross entropy loss
function with a softmax in the final layer. The lattice-based convolutional architecture
is summarized in Figure~\ref{fig:architecture}.
The networks are trained using the
Adam gradient algorithm with learning rate $10^{-4}$ for 300 epochs. We reserve
10\% of the data for testing. Results are shown in Figure~\ref{fig:comparison}.

% The lattice-based convolution and classical convolution networks perform
% similarly on the training set. However, the lattice convolution network's
% predictions generalize more accurately to the testing set ($\approx 7\%$ improvement in testing accuracy).
% This improvement is  at the expense of more variability in performance during
% training. (Our code is available at \texttt{https://github.com/alfred-tarski/latticeconv} .)

\section{Discussion}

Our proposed featurization for persistence modules is rather naive, but well
adapted to the use of lattice convolutions as a data processing method. The
lattice convolutional neural network shows promise as a method for classifying
features arising from a multiparameter persistence module. The algebraic
perspective on partially ordered sets, exemplified by lattices, may also offer
approaches to featurizing more complex invariants of persistence modules. In
particular, the incidence algebra may offer a natural way to represent the rank
invariant~\cite{carlsson_theory_2009} in a way amenable to convolution-like
operations.
We hope with these brief experiments to inspire further work on featurizing
multidimensional persistence for use in machine learning algorithms.

\begin{figure}[h!]
  \caption{The architecture of our lattice-based convolutional neural network.}\label{fig:architecture}
    \begin{center}
    \includegraphics[width=0.75\textwidth]{nn.jpg}
    \end{center}
\end{figure}

\begin{figure}[h!]
  \caption{Comparison learning curves for the lattice neural network and
    standard convolutional neural network.}\label{fig:comparison}
    \begin{center}
    %\includegraphics[width=0.75\textwidth]{training.png}
    \includegraphics[width=0.9\textwidth]{testing.png}
    \end{center}
\end{figure}

\section*{Acknowledgments}
The authors would like to thank Chris Wendler for replicating our original experiment. We would like to thank the reviewers for providing helpful feedback. The author [HR] is partially supported by Office of Naval Research (Grant No. N00014-1442-16-1-2010).
\small
\bibliographystyle{plain}
\bibliography{latticenn}
